---
title: "dgevbhm_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dgevbhm_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we present an example workflow to apply our R code for fitting the duration-dependent Generalized Extreme Value distribution (dGEV) in a Bayesian Hierarchical framework (Rischmuller et al., 2026). 
We use sub-daily precipitation records from the Global Sub-Daily Rainfall dataset (Lewis et al., 2019) over the United Kingdom.

```{r setup}
library(dgevbhm)
```

We use the *data_read* function to read the sub-daily precipitation data directly from the directory. The user can indicate, for which durations the dGEV models will be fitted.

```{r data_path, eval=FALSE}
# This is the path of the data in our work PC. Please use your own!
UK_data_full = data_read(dir_path = "../../UK/",
                         missing_data = 5,
                         durs = c(1,3,6,12,24,48),
                         cores = 4)
```


