---
title: " `dgevbhm` vignette "
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dgevbhm_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we present an example workflow to apply our R code for fitting the duration-dependent Generalized Extreme Value distribution (dGEV) in a Bayesian Hierarchical framework (Rischmuller et al., 2026). 
We use sub-daily precipitation records from the Global Sub-Daily Rainfall dataset (Lewis et al., 2019) over the United Kingdom.

```{r setup}
library(dgevbhm)
```

The `data_read()` function reads in sub-daily precipitation data directly from a directory `dir_path` of text files.
`missing_data` specifies the minimum of percentage of missing data per year, otherwise the year is disregarded from the data.
`durs` are the durations of the precipitation maxima for IDF/dGEV fitting.
`cores` allows 

```{r data_path, eval=FALSE}
UK_data_full = data_read(dir_path = "../../UK/",
                         missing_data = 5,
                         durs = c(1,3,6,12,24,48),
                         cores = 4)
```

```{r full_data_length}
length(UK_data_full$y)
```

Out of 1903 files 47 stations were selected with requirements mentioned above.

## Selecting the rain gauges

`dgev_bhm()` is the central function of our package. It is able to fit the Bayesian Hierarchical dGEV models for a chosen sample of rain gauges. There, the user can choose, if the shape parameter of the dGEV is fixed over durations (resulting in the same shape parameter for all durations, $xi_d$) or fixed over space (resulting in the same shape parameter for all rain gauges, $xi_j$).

```{r mean_median}
# mean and median hourly precipitation rate (mm)
mean(UK_data_full$amp)
median(UK_data_full$amp)
```

Below we show an example of selecting similar stations using annual mean precipitation `amp`:

```{r selection}
med_amp <- median(UK_data_full$amp)
# similar stations: selection of 25 stations that are closest to the median precipitation rate
sim_stations <- which(rank(abs(UK_data_full$amp - med_amp)) <= 25)
# plotting a histogram of the 25 precipitation rates
hist(UK_data_full$amp[sim_stations], main = 'histogram', xlab = 'stations')
```

```{r data_names}
names(UK_data_full) #checking header of the GSDR data
```
```{r latlondim}
dim(UK_data_full$latlon) #dimension of the latlon 
```

```{r clean_data}
# clean the data: only select the 25 rain gauges for the further example
UK_data_full$y <- UK_data_full$y[c(sim_stations)]
UK_data_full$amp <- UK_data_full$amp[sim_stations]
UK_data_full$latlon <- UK_data_full$latlon[sim_stations,]
UK_data_full$stationID <- UK_data_full$stationID[sim_stations]
```

## Fit the Bayesian hierarchical models

```{r dgev_bhm_run_d, eval=FALSE}
# Fit the dGEV-BHM with the shape parameter fixed over durations (shp_d = 'd'). We opt for 2000 iterations of the ... , where the first half is used as burn-in. 
# We assume that 2 cores are available for computation.

# WARNING: May take some time. You can also load the prepared fitted models.
system.time({
    fit_xi_d <- dgev_bhm(data = UK_data, chains = 2, iter = 2000,cores = 2,shp_d = 'd')
})
```

```{r dgev_bhm_run_j, eval=FALSE}
# Fit the dGEV-BHM with the shape parameter fixed over space (shp_d = 'j'). We opt for 2000 iterations of the MCMC, where the first half is used as burn-in. 
# We assume that 2 cores are available for computation.

# WARNING: May take some time. You can also load the prepared fitted models.
system.time({
    fit_xi_j <-  dgev_bhm(data = UK_data, chains = 2, iter = 2000,cores = 2,shp_d = 'j')
})
```

```{r fit_names}
#structure of the fitted models
names(fit_xi_d) 
names(fit_xi_j)
```

